ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-16 notebooks/poc.qmd
new illustrative chart

@@

scale = 0.1
for (x, y) in inputs
    push!(u, scale * gradient(f1, [x, y][:,:])[1][1])
    push!(v, scale * gradient(f1, [x, y][:,:])[1][2])
end

>>>>

```{julia}
using Colors
col_pal = palette(:seaborn_colorblind)
Random.seed!(1234)
using CounterfactualExplanations.Generators: ∇

λ₁ = 0.1
λ₂ = 1.0
λ₃ = 0.5
Λ = [λ₁, λ₂, λ₃]
η = 0.01

opt = Flux.Optimise.Descent(η)

generator_dict = OrderedDict(
    "Wachter" => WachterGenerator(λ = 0.3, opt=opt),
    "ECCCo (no EBM)" => ECCCoGenerator(λ = [λ₁,λ₂,0.0], opt=opt),
    "ECCCo (no CP)" => ECCCoGenerator(λ = [λ₁,0.0,λ₃], opt=opt),
    "ECCCo" => ECCCoGenerator(λ = Λ, opt=opt),
)

# Gradient field:
function loss_grads(generator, model, ce, x)
    x = Float32.(x)
    _ce = deepcopy(ce)
    _ce.s′ = x
    return ∇(generator,M,_ce)
end

meshgrid(x, y) = (repeat(x, outer=length(y)), repeat(y, inner=length(x)))

xlims, ylims = extrema(X, dims=2)
xrange = range(xlims..., length=10)
yrange = range(ylims..., length=10)
x1, x2 = meshgrid(xrange, yrange)
inputs = zip(x1, x2)

# Plot:
ces = Dict{Any,Any}()
plts = []
for (name, generator) in generator_dict

    # CE:
    ce = generate_counterfactual(
        x_factual, target, counterfactual_data, M, generator;
        initialization=:identity, 
        converge_when=:generator_conditions,
    )

    # Main plot (path):
    plt = Plots.plot(
        ce, title=name, alpha=0.1, cbar=false, 
        axis=nothing, length_out=10, contour_alpha=0.0,
        legend = false,
        palette = col_pal,
    )

    # Generated samples:
    if name ∈ ["ECCCo","ECCCo (no CP)"]
        _X = distance_from_energy(ce, return_conditionals=true)
        Plots.scatter!(
            _X[1,:],_X[2,:], color=col_pal[end-1], shape=:star5, 
            ms=10, label="x̂|$target", alpha=0.1
        )
    end

    # Gradient field:
    u = []
    v = []
    for (x, y) in inputs
        g = -loss_grads(generator, M, ce, [x, y][:,:])
        push!(u, η * g[1])
        push!(v, η * g[2])
    end
    Plots.quiver!(x1, x2, quiver=(u, v), color=col_pal[5])

    push!(plts, plt)
    ces[name] = ce
end
plt = Plots.plot(plts..., size=(500,520))
display(plt)
savefig(plt, joinpath(output_images_path, "poc_gradient_fields.png"))
```

@@

for (x, y) in inputs
    push!(u, scale * gradient(f1, [x, y][:,:])[1][1])
    push!(v, scale * gradient(f1, [x, y][:,:])[1][2])
end

### 2023-05-16 paper/paper.tex
new illustrative chart

We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms. We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data. Since these data are generated by distributions of varying degrees of complexity, they allow us to gauge how complexity affects the different generators.

>>>>

We use both synthetic and real-world datasets from different domains, all of which are publically available and commonly used to train and benchmark classification algorithms. We synthetically generate a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the well-known \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data. Since these data are generated by distributions of varying degrees of complexity, they allow us to assess how the generators and our proposed evaluation metrics handle this.
